Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-08-31 paper.Rnw
last edits - ready to go out!

We apply the \kge{trained} random forest \kge{model} from \citet{hare:2016} to another set of bullets from the Hamby study with \Sexpr{dim(bullet_test)[1]} rows of land comparisons. We first consider a global visualization of the relationship between the random forest predictions and the model features with a parallel coordinate plot of the training data (top facet) and testing data (bottom facets) predictions (\autoref{fig:bullet-plot-rfpcp}). The majority of same source observations with random forest probabilities close to 1 have a clear pattern of corresponding feature values. The known same source observations where the random forest is wrong (does not return probabilities close to 1) have feature values that reflect those of observations where same source is known to be false.

@@

To try to identify a set of LIME tuning parameter values with the most trustworthy set of explanations, an assessment metric plot is considered (\autoref{fig:bullet-assess-metric}). All density simulation methods performed well based on the three metrics. This is an interesting result, since \autoref{fig:bullet-feature-heatmap} shows that the density methods results in global explanations. The metric values for the bin based methods do not agree across metrics. For example, 2 and 3 quantile bins performed well according to the average $R^2$ but poorly according to MSEE. As a result, with the bin based methods, it is difficult to know which method to recommend. The applications using a power of 0.5 perform the best or as well as the other powers across all simulation methods suggesting that the power that leads to a more global explanation is preferred by LIME. Perhaps we can say that the kernel density method \kge{or} 3 equal bins \sout{, or 3 quantile bins} are \kge{the best} \sout{some of the better} performing methods, but unfortunately, \autoref{fig:bullet-assess-metric} leaves us without a clear indication of which set of explanations to use to explain the random forest.

\kge{To provide a more detailed view of explanations obtained using different tuning parameter values,} \sout{To better understand the explanations provided by these three methods,} we take a closer look at explanations \sout{from these methods} for two observations of interest, which will be referred to as case M (a known match) and case NM (a known non-match)\kge{, using explanation scatterplots}. Figure \ref{fig:bullet-eois} includes the visual representation plots of explanations (from \emph{lime}) and explanation scatterplots for cases M and NM. Plots for 3 equal bins and 3 quantile bins are included \kge{to depict tuning parameter values with good (3 equal bins) and mediocre (3 quantile bins) performance based on the assessment metric plot}. Plots for the kernel density simulation method are included in Appendix~\ref{exp-scatter_plus} (Figure \ref{fig:bullet-expl-scat-density}). \sout{The scatterplots provide several insights into the LIME explanations. Here, we focus on the explanation for case NM with 3 quantile bins, but similar statements can be made about the other plots.}

\kge{First, focus on the explanation for case NM with 3 quantile bins.} The explanation plot from \emph{lime} \sout{for case NM with 3 quantile bins} shows that an observed CCF value greater than 0.320 supports a random forest prediction in favor of a match, and the scatter plots show that many of the simulated values with CCF greater then 0.320 have random forest predictions greater than 0.5. Additionally, the LIME explanation indicates that a value of matches less than or equal to 1.66 contradicts a prediction in favor of a match, and the scatter plots show that almost all simulated values with a value of matches less than or equal to 1.66 have random forest predictions close to 0. While the LIME explanation makes sense based on the observations made from the explanation scatter plot, the plot also indicates that LIME falls short of providing a good explanation of the random forest prediction for case NM. Based on the scatter plot of CCF versus matches, a better explanation would be that because case NM has a value of CCF less than 0.8 and a value of matches less than 7, the random forest provides a prediction supporting a non-match. The relationship between CCF and rough correlation does not provide much evidence to support the random forest prediction one way or the other, but there is a pentagon shaped region at the bottom of the scatter plot of matches versus rough correlation with mostly predictions close to 0 that case NM falls in and supports the random forest prediction of a non-match.

\kge{The other explanation scatterplots can be interpreted in a similar manner. Overall, the plots identify that the 3 quantile bin explanations are not faithful to the random forest model predictions. The 3 equal bin explanations are more faithful to the random forest, but the regions captured by the bins still do not align with the regions with similar probabilities produced by the random forest.}

@@

For $x_e$ and \kge{a} set of tuning parameter values $t$, let $\textbf{X}_{e,t}'$ be the LIME simulated dataset with $K$ features and $S$ observations such that $x'_{e,t,s}$ is the feature vector for simulated data observation $s$ \kge{corresponding to observation/explanation $e$ and tuning parameter values $t$}. Let $\textbf{Z}'_{e,t}$ be the matrix of \kge{simulated data transformed to bin indicator variables (for bin based simulation methods) or standardization (for density based simulation methods)} \sout{interpretability transformed simulated data} with $K$ features and $S$ observations such that $z'_{e,t,s}$ is the interpretability transformed feature vector for observation/explanation $e$, tuning parameter values $t$, and simulated data observation $s$. Note that $z_{e,t}$ will represent the \sout{interpretability} transformed version of $x_e$.

Next, let $\omega_t$ represent a proximity distance metric \kge{corresponding to tuning parameter values $t$}. Then $\omega_t\left(x_e, x'_{e,t,s}\right)$ is the weight assigned to $x'_{e,t,s}$, which is the proximity between $x_e$ and $x'_{e,t,s}$. Allow $g_{e,t}$ to be the explainer model for an explanation $e$ and tuning parameter values $t$. Thus, $g_{e,t}\left(z'_{e,s,t}\right)$ is the explainer model prediction for the interpretability transformed simulated data observation $s$.

@@

\noindent \kge{where $\mathcal{L}$ was originally defined in \citet{ribeiro:2016}.}\\
\\
\emph{Mean squared explanation error} is denoted by MSEE and computed as

>>>>

We apply the trained random forest model from \citet{hare:2016} to another set of bullets from the Hamby study with \Sexpr{dim(bullet_test)[1]} rows of land comparisons. We first consider a global visualization of the relationship between the random forest predictions and the model features with a parallel coordinate plot of the training data (top facet) and testing data (bottom facets) predictions (\autoref{fig:bullet-plot-rfpcp}). The majority of same source observations with random forest probabilities close to 1 have a clear pattern of corresponding feature values. The known same source observations where the random forest is wrong (does not return probabilities close to 1) have feature values that reflect those of observations where same source is known to be false.

@@

To try to identify a set of LIME tuning parameter values with the most trustworthy set of explanations, an assessment metric plot is considered (\autoref{fig:bullet-assess-metric}). All density simulation methods performed well based on the three metrics. This is an interesting result, since \autoref{fig:bullet-feature-heatmap} shows that the density methods results in global explanations. The metric values for the bin based methods do not agree across metrics. For example, 2 and 3 quantile bins performed well according to the average $R^2$ but poorly according to MSEE. As a result, with the bin based methods, it is difficult to know which method to recommend. The applications using a power of 0.5 perform the best or as well as the other powers across all simulation methods suggesting that the power that leads to a more global explanation is preferred by LIME. Perhaps we can say that the kernel density method or 3 equal bins  are the best performing methods, but unfortunately, \autoref{fig:bullet-assess-metric} leaves us without a clear indication of which set of explanations to use to explain the random forest.

To provide a more detailed view of explanations obtained using different tuning parameter values, we take a closer look at explanations  for two observations of interest, which will be referred to as case M (a known match) and case NM (a known non-match), using explanation scatterplots. Figure~\ref{fig:bullet-eois} includes the visual representation plots of explanations (from \emph{lime}) and explanation scatterplots for cases M and NM. Plots for 3 equal bins and 3 quantile bins are included to depict tuning parameter values with good (3 equal bins) and mediocre (3 quantile bins) performance based on the assessment metric plot. Plots for the kernel density simulation method are included in Appendix~\ref{exp-scatter_plus} (Figure \ref{fig:bullet-expl-scat-density}).

First, focus on the explanation for case NM with 3 quantile bins. The explanation plot from \emph{lime} \sout{for case NM with 3 quantile bins} shows that an observed CCF value greater than 0.320 supports a random forest prediction in favor of a match, and the scatter plots show that many of the simulated values with CCF greater then 0.320 have random forest predictions greater than 0.5. Additionally, the LIME explanation indicates that a value of matches less than or equal to 1.66 contradicts a prediction in favor of a match, and the scatter plots show that almost all simulated values with a value of matches less than or equal to 1.66 have random forest predictions close to 0. While the LIME explanation makes sense based on the observations made from the explanation scatter plot, the plot also indicates that LIME falls short of providing a good explanation of the random forest prediction for case NM. Based on the scatter plot of CCF versus matches, a better explanation would be that because case NM has a value of CCF less than 0.8 and a value of matches less than 7, the random forest provides a prediction supporting a non-match. The relationship between CCF and rough correlation does not provide much evidence to support the random forest prediction one way or the other, but there is a pentagon shaped region at the bottom of the scatter plot of matches versus rough correlation with mostly predictions close to 0 that case NM falls in and supports the random forest prediction of a non-match.

The other explanation scatterplots can be interpreted in a similar manner. Overall, the plots identify that the 3 quantile bin explanations are not faithful to the random forest model predictions. The 3 equal bin explanations are more faithful to the random forest, but the regions captured by the bins still do not align with the regions with similar probabilities produced by the random forest.

@@

For $x_e$ and a set of tuning parameter values $t$, let $\textbf{X}_{e,t}'$ be the LIME simulated dataset with $K$ features and $S$ observations such that $x'_{e,t,s}$ is the feature vector for simulated data observation $s$ corresponding to observation/explanation $e$ and tuning parameter values $t$. Let $\textbf{Z}'_{e,t}$ be the matrix of simulated data transformed to bin indicator variables (for bin based simulation methods) or standardization (for density based simulation methods)  with $K$ features and $S$ observations such that $z'_{e,t,s}$ is the interpretability transformed feature vector for observation/explanation $e$, tuning parameter values $t$, and simulated data observation $s$. Note that $z_{e,t}$ will represent the  transformed version of $x_e$.

Next, let $\omega_t$ represent a proximity distance metric corresponding to tuning parameter values $t$. Then $\omega_t\left(x_e, x'_{e,t,s}\right)$ is the weight assigned to $x'_{e,t,s}$, which is the proximity between $x_e$ and $x'_{e,t,s}$. Allow $g_{e,t}$ to be the explainer model for an explanation $e$ and tuning parameter values $t$. Thus, $g_{e,t}\left(z'_{e,s,t}\right)$ is the explainer model prediction for the interpretability transformed simulated data observation $s$.

@@

\noindent where $\mathcal{L}$ was originally defined in \citet{ribeiro:2016}.\\
\\
\emph{Mean squared explanation error} is denoted by MSEE and computed as
